---
title: "The classic SESOI power analysis, and what it cannot tell you"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The classic SESOI power analysis, and what it cannot tell you}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## The classic recipe

The most common SESOI power analysis powers a *null-hypothesis test* at a
true effect equal to the SESOI: "if the effect is exactly as small as the
smallest thing I care about, I still detect it with 80% probability". In
`powerape` this is `claim = "detect"` with the target set to the SESOI:

```{r}
library(powerape)

d <- ape_dgp(model = "probit",
             focal = pa_var("treat", "binary", p = 0.5),
             baseline = 0.30)
d <- set_ape(d, target = 0.05)   # truth = SESOI = 5 pp

ape_power(d, n = 5000, claim = "detect", sesoi = 0.05,
          nsim = 600, seed = 1)
```

Note what the outcome distribution shows: even with ample detection power,
essentially **no study can claim the effect is meaningfully large**
(`minimum` outcome near zero), because when the truth sits at the SESOI the
CI lower bound almost never clears it. Detection is all the classic recipe
buys.

## The boundary is a wall, not a target

Asking for minimum-effect power *at* the SESOI is statistically degenerate,
and `powerape` refuses with an explanation rather than returning a useless
number:

```{r, error = TRUE}
ape_power(d, n = 5000, claim = "minimum", sesoi = 0.05)
```

On the null boundary of the minimum-effect test, the claim succeeds with
probability equal to the test size (about 2.5% with a 95% CI) at *any*
sample size.

## The two coherent moves

Either stay with detection at truth = SESOI (this vignette), or assume a
planning value **above** the SESOI and power the minimum-effect claim.
The second buys a stronger conclusion and costs more observations:

```{r}
## detection power for truth = SESOI = 0.05
ape_n(d, power = 0.80, claim = "detect", nsim = 500, seed = 2)

## minimum-effect power for truth = 0.08 against SESOI = 0.05
d8 <- set_ape(d, target = 0.08)
ape_n(d8, power = 0.80, claim = "minimum", sesoi = 0.05,
      nsim = 500, seed = 3)
```

The gap between those two n's is the price of concluding "meaningfully
large" instead of "not zero". Making that price visible -- in APE units,
before data collection -- is precisely what a SESOI-based design discussion
needs.

## Size audit

A useful sanity check of the whole simulation pipeline: at a true APE of
zero, directional detection should fire at half the nominal two-sided rate.

```{r}
d0 <- set_ape(d, target = 0)
ape_power(d0, n = 2000, claim = "detect", nsim = 600, seed = 4)
```
