---
title: "Powering for equivalence: showing an effect is too small to matter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Powering for equivalence: showing an effect is too small to matter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6, fig.height = 4)
```

## The claim

Sometimes the research goal is to show an effect is *negligible*: a cheaper
treatment is not meaningfully worse, a feared side effect is not
meaningfully present. The equivalence claim succeeds when the whole
confidence interval lies inside (-SESOI, +SESOI). A non-significant test
is **not** evidence of absence; a CI inside the equivalence bounds is.

```{r}
library(powerape)

d <- ape_dgp(model = "probit",
             focal = pa_var("treat", "binary", p = 0.5),
             covariates = list(pa_var("age", "normal", mean = 45, sd = 12)),
             baseline = 0.30, signal = 0.10)
d <- set_ape(d, target = 0)    # planning value: no effect at all
```

The coherence guard requires |target| < SESOI: you cannot power for
equivalence while assuming a true effect on or beyond the bounds.

## Power and required n

```{r}
ape_power(d, n = 2000, claim = "equivalence", sesoi = 0.05,
          nsim = 600, seed = 1)
```

```{r}
ape_n(d, power = 0.80, claim = "equivalence", sesoi = 0.05,
      nsim = 500, seed = 2)
```

Equivalence needs the CI to be *narrow*, not just centered near zero, so
sample sizes are driven entirely by the SESOI: halving the SESOI roughly
quadruples the required n.

```{r}
cv <- ape_curve(d, n = seq(500, 3500, by = 750), claim = "equivalence",
                sesoi = 0.05, nsim = 300, seed = 3)
plot(cv, target_power = 0.80)
```

## A robustness note

The width of the CI for an APE depends on the baseline rate (binomial
variance peaks at 0.5), so equivalence power inherits that dependence.
`ape_robust()` makes it visible:

```{r}
ape_robust(d, n = 2600, claim = "equivalence", sesoi = 0.05,
           vary = list(baseline = c(0.20, 0.50)), grid_points = 3,
           nsim = 300, seed = 4, nmax = FALSE)
```

## Reporting

`power_statement()` works for equivalence designs too, and states the CI
convention explicitly -- with a 95% interval the implied two one-sided
tests run at the conservative 2.5% level (Riesthuis, 2024).

```{r}
pw <- ape_power(d, n = 2600, claim = "equivalence", sesoi = 0.05,
                nsim = 500, seed = 5)
power_statement(pw)
```
