---
title: "Power for average interaction effects (AIEs)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Power for average interaction effects (AIEs)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## The estimand

"Does the treatment work differently for women?" is a question about the
**average interaction effect (AIE)** -- in a nonlinear model *not* the
interaction coefficient (Ai & Norton, 2003), but the averaged
double difference / cross-partial that the
[`ginteff`](https://github.com/jespernwulff/ginteff) package estimates.
`powerape` reimplements the AIE internally for speed, unit-tested to
reproduce `ginteff`'s estimates and delta-method standard errors exactly.

## Specifying and pinning an interaction design

Add a `moderator` to the DGP; the index then includes both mains and the
interaction. Pinning happens in effect-size units via `set_aie()`, which
takes three explicit anchors:

* `main_focal`, `main_moderator` -- the *conditional-at-reference APEs*:
  the APE of each variable with the other held at its reference value
  (0 for binary variables);
* `target` -- the AIE itself.

For a binary-by-binary design this is exactly a bijection with the four
counterfactual cell rates: baseline p00, p10 = p00 + main_focal,
p01 = p00 + main_moderator, p11 = p00 + main_focal + main_moderator + AIE.

```{r}
library(powerape)

d <- ape_dgp(model = "probit",
             focal = pa_var("treat", "binary", p = 0.5),
             moderator = pa_var("female", "binary", p = 0.55),
             covariates = list(pa_var("age", "normal", mean = 45, sd = 12)),
             baseline = 0.30, signal = 0.10)
d <- set_aie(d, target = 0.08, main_focal = 0.10, main_moderator = 0.03)
d
```

Feasibility is checked at every step (all four cell rates must stay inside
(0, 1); with continuous variables the AIE plateaus in the interaction
coefficient), with errors reporting the attainable range.

## Interactions are expensive

The estimand rides on the DGP, so the usual functions power the AIE
directly:

```{r}
ape_power(d, n = 3000, claim = "detect", nsim = 500, seed = 1)
```

Compare that with how easy an equally sized *main effect* would be:

```{r}
d_main <- ape_dgp(model = "probit",
                  focal = pa_var("treat", "binary", p = 0.5),
                  covariates = list(pa_var("age", "normal", mean = 45, sd = 12)),
                  baseline = 0.30, signal = 0.10)
d_main <- set_ape(d_main, target = 0.08)
ape_power(d_main, n = 3000, claim = "detect", nsim = 500, seed = 2)
```

Same 8-point effect, same n -- a fraction of the power. The folk theorem
that interaction studies need several times the sample is here made exact,
in the units researchers actually interpret, via `ape_n()`:

```{r}
ape_n(d, power = 0.80, claim = "detect", nsim = 500, seed = 3)
```

All claims work for AIEs: power a *meaningful* interaction with
`claim = "minimum"` and an AIE SESOI, or show treatment-effect homogeneity
with `claim = "equivalence"`.

## Correspondence with ginteff (and Stata)

powerape's binary AIE is the 0-to-1 counterfactual contrast. In `ginteff`
that corresponds to **factor** variables passed to `dydxs` (the analog of
Stata's `i.` factor syntax): discrete change from the base level. Numeric
variables in `firstdiff` instead get "+1 unit from the observed value" --
for a 0/1 dummy that extrapolates treated units to 2, a different estimand.
When validating a powerape design against `ginteff` on pilot data, declare
binary variables as factors.

## References

* Ai, C., & Norton, E. C. (2003). Interaction terms in logit and probit
  models. *Economics Letters*, 80(1), 123-129.
* Radean, M. (2023). ginteff: A generalized command for computing
  interaction effects. *The Stata Journal*, 23(2).
