---
title: "Justifying the error rate: alpha by sample size, or by error costs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Justifying the error rate: alpha by sample size, or by error costs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6, fig.height = 4)
has_alphaN <- requireNamespace("alphaN", quietly = TRUE)
```

## The default, and the argument

Every claim in powerape is tested at `alpha = 0.05` in the form its
literature uses: detection is the two-sided test read off the 95%
interval, the minimum-effect claim a one-sided test and equivalence a
TOST, both read off the 90% interval. The level is an argument of every
power function, so a stricter convention is one keyword away:

```{r}
library(powerape)

d <- ape_dgp(model = "probit",
             focal = pa_var("program", "binary", p = 0.5),
             covariates = list(pa_var("z", "normal")),
             baseline = 0.30, signal = 0.10)
d <- set_ape(d, target = 0.10)

ape_power(d, n = 2000, claim = "minimum", sesoi = 0.05, nsim = 400, seed = 1)
ape_power(d, n = 2000, claim = "minimum", sesoi = 0.05, alpha = 0.01,
          nsim = 400, seed = 1)
```

Maier and Lakens (2022) give two reasons not to stop at the default.
First, with a fixed budget of observations the error rate that
minimizes the combined cost of Type I and Type II errors is rarely
exactly .05. Second, in large samples a *p* value just below .05 can be
more likely under the null than under the alternative (Lindley's
paradox), so the level should fall as *n* grows. powerape supports
both justifications.

## Alpha as a function of the sample size

`alpha` may be a function of *n*. It is evaluated at every candidate
sample size, so `ape_n()` designs jointly over the pair (alpha(*n*),
*n*), and `ape_curve()`, `ape_mde()`, and `ape_robust()` use the level
that belongs to their *n*. Any rule works; the natural one for
regression coefficients is the Bayes-factor calibration of Wulff and
Taylor (2024) in the alphaN package, where `BF = 1` is the largest
level that avoids Lindley's paradox and `BF = 3` asks a significant
result to carry moderate evidence.

```{r, eval = has_alphaN}
rule <- function(n) alphaN::alphaN(n, BF = 3)
rule(c(500, 2000, 8000))

ape_n(d, power = 0.80, claim = "minimum", sesoi = 0.05, alpha = rule,
      nsim = 300, seed = 2, confirm = FALSE)
```

The requirement rises well above the fixed-.05 answer (about 2,100 for
this design), because the level keeps tightening as the search moves
to larger *n*. The returned object stores the rule and the level it
produced at the chosen *n*, and `power_statement()` says so.

```{r, eval = has_alphaN}
pw_rule <- ape_power(d, n = 5000, claim = "minimum", sesoi = 0.05,
                     alpha = rule, nsim = 400, seed = 3)
pw_rule
power_statement(pw_rule)
```

## Alpha from error costs

The second route is decision-theoretic. For a design with a fixed *n*,
the weighted combined error rate

$$ w(\alpha) = \frac{cost \cdot \alpha + prior \cdot \beta(\alpha)}{cost + prior} $$

weighs a Type I error `cost` times as heavily as a Type II error and
the two hypotheses by their prior odds (Mudge et al., 2012). powerape
evaluates the whole curve from **one** simulation: `ape_power()` keeps
the estimate and standard error of every replication, and power at any
level is a re-threshold of those draws. `power_at()` exposes the curve;
`ape_alpha()` minimizes or balances it.

```{r}
pw <- ape_power(d, n = 2000, claim = "minimum", sesoi = 0.05,
                nsim = 1000, seed = 4)
power_at(pw, alpha = c(0.005, 0.01, 0.025, 0.05, 0.10))

ja <- ape_alpha(pw, cost = 4, prior = 1)      # Cohen's 4:1 weighting
ja
plot(ja)
```

Two things to notice. The objective is flat around its optimum, so the
result reports the range of levels that cannot be told apart from the
optimum at this number of replications; that range, not the point, is
what to preregister unless `nsim` is large. And the search is capped at
.05: when the unconstrained optimum lies above the conventional level,
as it does for designs below their required *n* with equal error
costs, the function says so rather than returning it, following Maier
and Lakens's advice that raising alpha needs a justification of its own
and that journals may accept only lowering.

```{r}
ape_alpha(pw, cost = 1, prior = 1)
```

Balancing instead of minimizing asks for `cost * alpha = prior * beta`;
at a design's required *n* for 80% power, Cohen's 4:1 weighting returns
the familiar .05/.20 pair by construction.

## The frontier level: one alpha for a whole MDE grid

`ape_alpha()` needs a pinned effect, because its Type II error is one
minus power *at the planning value*. In a minimum-detectable-effect
analysis the effect is the answer, not an input, and re-optimizing the
level at every candidate effect chases a moving target. The way out is
a fixed point: the level that is optimal *for the effect that is just
detectable at the target power*. Under the normal approximation that
first-order condition loses both *n* and the standard error, so one
level prices every sample size, and `alpha_frontier()` returns it in
closed form.

```{r}
alpha_frontier(cost = 4, power = 0.80)                    # two-sided, 0.0274
alpha_frontier(cost = 1, power = 0.80, claim = "minimum") # exactly 1 - power
```

Passing it as `alpha` re-prices an MDE analysis (or a grid of them)
under the justified level:

```{r, eval = FALSE}
ape_mde(d, n = 3000, claim = "detect",
        alpha = alpha_frontier(cost = 4, power = 0.80), seed = 6)
```

At equal costs the one-sided frontier level is exactly `1 - power` --
the frontier is where alpha and beta balance. Unlike `ape_alpha()`,
the frontier level is an approximation (normal, and pinned to the
just-detectable effect) and carries no .05 cap, so treat levels above
the convention with the same caution as an uncapped `ape_alpha()`
optimum.

## Checking the size, and reporting

Because the Type I error of a SESOI claim is its size at the claim
boundary, `size = TRUE` simulates the world in which the true effect
equals the SESOI and reports the realized error rate at the chosen
level, something analytic compromise-power tools cannot do. On the
standard routes it matches the nominal level within Monte Carlo error;
on clustered panels with few units it may not, and then the realized
number is the one to report.

```{r}
ja4 <- ape_alpha(pw, cost = 4, prior = 1, size = TRUE, nsim_size = 500,
                 seed = 5)
ja4
power_statement(ja4)
```

The statement names the objective, the weights, the level, the flat
range, the cap, and the verification, ready for a preregistration.

## References

Maier, M., & Lakens, D. (2022). Justify your alpha: A primer on two
practical approaches. *Advances in Methods and Practices in
Psychological Science, 5*(2).

Mudge, J. F., Baker, L. F., Edge, C. B., & Houlahan, J. E. (2012).
Setting an optimal alpha that minimizes errors in null hypothesis
significance tests. *PLOS ONE, 7*(2), e32734.

Wulff, J. N., & Taylor, L. (2024). How and why alpha should depend on
sample size: A Bayesian-frequentist compromise for significance
testing. *Strategic Organization, 22*(3), 550--581.
