---
title: "Powering the minimum-effect claim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Powering the minimum-effect claim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6, fig.height = 4)
```

## Why power for an APE at all?

In a probit or logit model, the coefficient is not the quantity anyone
interprets. Applied conclusions are stated in probability points -- "the
treatment raises take-up by 10 percentage points" -- which is the **average
partial effect (APE)**. The APE depends on *all* coefficients and on the
covariate distribution, so power tools that work in coefficient or Cohen's-d
units cannot answer the question practitioners actually have.

`powerape` lets you specify the target in APE units and simulates the whole
pipeline: draw covariates, draw outcomes, fit the model by maximum
likelihood, compute the APE with a delta-method Wald confidence interval,
and evaluate a **claim** on that interval (Riesthuis, 2024):

* `"minimum"` -- the CI lower bound exceeds the smallest effect size of
  interest (SESOI): the effect is *meaningfully* large. This is the default
  claim and the package's reason to exist.
* `"detect"` -- the CI excludes zero (directional).
* `"equivalence"` -- the CI lies inside (-SESOI, +SESOI): the effect is too
  small to matter.

## Step 1: describe the world

```{r}
library(powerape)

d <- ape_dgp(
  model       = "probit",
  focal       = pa_var("treat", "binary", p = 0.5),
  covariates  = list(pa_var("age", "normal", mean = 45, sd = 12),
                     pa_var("female", "binary", p = 0.55)),
  correlation = 0.2,
  baseline    = 0.30,   # P(Y = 1) with treat = 0
  signal      = 0.15    # latent pseudo-R2 of the covariates
)
```

The nuisance side is anchored on a friendly scale: the **baseline outcome
rate** and the covariates' **latent pseudo-R-squared**. Explicit
coefficients (`gamma`), pilot data (`ape_dgp_empirical()`), or a fitted
pilot model (`ape_dgp_from_fit()`) are the expert routes.

## Step 2: pin the true APE (the inversion step)

Suppose theory or prior evidence suggests a true effect of 10 percentage
points, and effects below 5 points would not justify the intervention: the
planning value is 0.10 and the SESOI is 0.05.

```{r}
d <- set_ape(d, target = 0.10)
d
```

`set_ape()` solves for the focal coefficient that makes the true APE
exactly 0.10 given everything else, and errors informatively if the target
is infeasible (an APE of +0.40 cannot exist at a baseline rate of 0.85).

## Step 3: power for the claim

```{r}
ape_power(d, n = 1500, claim = "minimum", sesoi = 0.05,
          nsim = 500, seed = 1)
```

Read the **outcome distribution**, not just the power number: some studies
would reject zero yet remain unable to claim the effect clears the SESOI
("detect-only") -- exactly the inconclusiveness Riesthuis (2024) warns
about.

## Step 4: required sample size and the power curve

```{r}
ape_n(d, power = 0.80, claim = "minimum", sesoi = 0.05,
      nsim = 500, seed = 2)
```

```{r}
cv <- ape_curve(d, n = seq(500, 3000, by = 500), claim = "minimum",
                sesoi = 0.05, nsim = 300, seed = 3)
plot(cv, target_power = 0.80)
```

## Step 5: how fragile is this to the contextual guesses?

The baseline rate and the covariate signal are guesses, and power computed
from a single guess is fragile (Hancock & Feng, 2025). `ape_robust()`
re-runs the analysis over a scenario grid and reports the worst case; with
`nmax = TRUE` it also finds **n_max**, the insurance-premium sample size
that holds the target power in the worst scenario.

```{r}
ape_robust(d, n = 2200, claim = "minimum", sesoi = 0.05,
           vary = list(baseline = c(0.20, 0.40)), grid_points = 3,
           nsim = 300, seed = 4, nmax = FALSE)
```

By default the APE is **re-pinned in every scenario** (`pin = "ape"`), so
only the precision channel varies; `pin = "coefficients"` instead holds the
coefficients and lets the implied effect drift.

## Step 6: the citable paragraph

```{r}
pw <- ape_power(d, n = 2200, claim = "minimum", sesoi = 0.05,
                nsim = 500, seed = 5)
power_statement(pw)
```

## References

* Hancock, G. R., & Feng, Y. (2025). nmax and the quest to restore caution,
  integrity, and practicality to the sample size planning process.
  *Psychological Methods*.
* Riesthuis, P. (2024). Simulation-based power analyses for the smallest
  effect size of interest: A confidence-interval approach for
  minimum-effect and equivalence testing. *AMPPS*, 7(2).
