---
title: "Panel designs: correlated random effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Panel designs: correlated random effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6, fig.height = 4)
```

## Why a panel module?

Much applied binary-outcome work is panel work: firm-years, person-waves,
plant-quarters. The workhorse model for that setting is a **correlated
random effects (CRE) probit or logit** -- pooled estimation with Mundlak
means and unit-clustered standard errors (Wooldridge, 2010, ch. 15). It
nests the pure random-effects model, allows the unit heterogeneity to
correlate with the regressors through their observed unit means, and is
what `xtprobit`-style advice converges to when the heterogeneity cannot
credibly be assumed independent.

The good news for design is that **the estimand does not change**. The
target is still the APE -- here the average-structural-function (ASF)
version: the effect on P(Y = 1) integrating over the unit effects, with
the Mundlak means held at their observed values. The pooled CRE estimator
recovers exactly that quantity even though its *coefficients* are
attenuated. Designing in APE units is right for panels too; that is the
point of the package.

What changes is the **accounting**: for panel DGPs, every sample-size
argument (`ape_power()`, `ape_curve()`, `ape_n()`, `ape_robust()`) counts
**units (clusters)**, not observations -- clusters are what drive power.

## Step 1: describe the panel world

```{r}
library(powerape)

d <- ape_dgp_panel(
  model      = "probit",
  focal      = pa_var("treat", "binary", p = 0.5, icc = 1),  # unit-level treatment
  covariates = list(pa_var("size", "normal", icc = 0.6)),
  n_periods  = 4,
  rho        = 0.30,
  cre_share  = 0.50,
  baseline   = 0.30,
  signal     = 0.10
)
d
```

Three knobs are panel-specific, and each is a quantity you can defend in a
grant application:

* `n_periods` -- balanced waves per unit.
* `rho` -- the share of latent variance coming from the unit effect,
  `Var(c) / (Var(c) + 1)`; the same convention `xtprobit` reports. Zero
  means no unobserved heterogeneity.
* `cre_share` -- how much of that unit effect is *correlated* with the
  regressors (loaded on the centered unit means of the time-varying
  variables); the remainder is a pure random effect. The estimator
  includes Mundlak means either way.

Per-variable persistence comes from `pa_var(..., icc = )`: `icc = 1` is a
time-constant, unit-level variable (a one-shot treatment, a sector);
values in (0, 1) mix unit-level and wave-level variation.

## Step 2: pin the APE

```{r}
d <- set_ape(d, target = 0.10)
```

Identical to the cross-sectional workflow -- the inversion solves the
focal coefficient so the true ASF-APE is exactly 0.10, integrating the
unit effects (in closed form for probit, by Gauss-Hermite quadrature for
logit), and errors informatively if the target is infeasible.

## Step 3: power, in units

```{r}
ape_power(d, n = 300, claim = "minimum", sesoi = 0.05,
          nsim = 400, seed = 1)
```

`n = 300` means 300 units observed for 4 periods -- 1,200 unit-period
observations, fitted by pooled probit with Mundlak means and
unit-clustered SEs in every replication. (Below 30 clusters the package
warns: cluster-robust inference is not trustworthy there.)

For archival panels the units are usually *given*, and the honest
question inverts: what is the smallest APE this panel can reliably
demonstrate? `ape_mde()` answers it with n still counted in units:

```{r}
ape_mde(d, n = 400, claim = "minimum", sesoi = 0.05,
        nsim = 400, seed = 9)
```

## Panels are not free observations

With a time-constant treatment, T waves of the same units add much less
than T-fold information: the outcome is correlated within units, so the
effective sample is closer to the number of clusters than to the number
of rows. Compare the panel above with a cross-section of the same *total*
size:

```{r}
cs <- ape_dgp(
  model      = "probit",
  focal      = pa_var("treat", "binary", p = 0.5),
  covariates = list(pa_var("size", "normal")),
  baseline   = 0.30,
  signal     = 0.10
)
cs <- set_ape(cs, target = 0.10)
ape_power(cs, n = 1200, claim = "minimum", sesoi = 0.05,
          nsim = 400, seed = 2)
```

How the two compare is a *computable property of the design*, not a rule
of thumb. Here — with half of the firm heterogeneity loading on
observable means that the Mundlak terms absorb — the panel's rows turn
out to be worth nearly as much as independent ones (high-replication
runs put the two powers within a point of each other). Push `rho` up, or
set `cre_share` toward zero so the heterogeneity is unobservable, and
the panel discount grows sharply. Reviewers who either count panel rows
as independent observations *or* apply a crude cluster-trial discount
are both guessing; this pair of calls replaces the guess.

Since `rho` is usually a guess, sweep it -- with `pin = "ape"` the true
effect is re-pinned in every scenario, so only the precision channel
moves:

```{r}
ape_robust(d, n = 300, claim = "minimum", sesoi = 0.05,
           vary = list(rho = c(0.1, 0.5)), grid_points = 3,
           nsim = 300, seed = 3, nmax = FALSE)
```

## Panel interactions: the AIE under CRE

Moderation questions survive the move to panels: does a governance reform
bite harder in public firms? The estimand is the **average interaction
effect (AIE)** -- the double difference of counterfactual cell
probabilities -- and for panel DGPs it is defined on the same ASF as the
APE: unit effects integrated out, Mundlak means held fixed. It is the
same population quantity a cross-sectional AIE design targets.

```{r}
di <- ape_dgp_panel(
  model      = "probit",
  focal      = pa_var("treat", "binary", p = 0.5, icc = 1),
  moderator  = pa_var("public", "binary", p = 0.4, icc = 0.7),
  covariates = list(pa_var("size", "normal", icc = 0.6)),
  n_periods  = 4,
  rho        = 0.30,
  cre_share  = 0.50,
  baseline   = 0.30,
  signal     = 0.10
)
di <- set_aie(di, target = 0.06, main_focal = 0.10, main_moderator = 0.05)
di
```

The three anchors are the familiar ones: the two conditional-at-reference
main-effect APEs plus the target AIE. One panel-specific detail is
handled for you: whenever the interaction term varies within units, the
estimating model includes the interaction's *own* Mundlak mean (the mean
of a product is not the product of the means).

Interactions are expensive, and panels make them more so -- the
interaction tax and the clustering penalty compound:

```{r}
ape_power(di, n = 400, claim = "detect", nsim = 250, seed = 4)
```

Even 400 units (1,600 observations) leave detection power low for a
6-point AIE. Budget accordingly -- `ape_n()` with a larger `nsim` gives
the required number of units, and `power_statement()` renders any of
these results as a citable methods paragraph that counts units and
periods correctly.

## What stands behind the numbers

The panel machinery is validated the same way as the rest of the package:
the ASF-based APE and the AIE double difference match Stata
(`probit ..., vce(cluster id)` + `margins`) to all printed digits on
fixed panels, with the small clustered-SE difference fully attributed to
Stata's observed-Hessian bread; the clustered sandwich matches
`sandwich::vcovCL` exactly; the AIE reproduces `ginteff` on the identical
fit; and the engine reduces to exact enumeration and to the Donner-Klar
cluster design effect in the cases where those are available. See the
validation battery in the repository for the full dossier.

Current scope: balanced panels, the parametric covariate route, and
binary-by-binary panel interactions; continuous panel pairs, unbalanced
panels, and time effects are on the roadmap.

## References

* Mundlak, Y. (1978). On the pooling of time series and cross section
  data. *Econometrica*, 46(1), 69-85.
* Riesthuis, P. (2024). Simulation-based power analyses for the smallest
  effect size of interest: A confidence-interval approach for
  minimum-effect and equivalence testing. *AMPPS*, 7(2).
* Wooldridge, J. M. (2010). *Econometric Analysis of Cross Section and
  Panel Data* (2nd ed.). MIT Press. Ch. 15.
